---
title: "BayesianLasso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BayesianLasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, message=FALSE, warning=FALSE}
library(BayesianLasso)

# locate package root
pkg_root <- normalizePath("..")

helpers_path <- file.path(pkg_root, "reproduction", "helpers")

helper_files <- list.files(helpers_path, pattern = "\\.R$", full.names = TRUE)

lapply(helper_files, source)
```

## Introduction

The \pkg{BayesianLasso} package provides efficient sampling algorithms for Bayesian Lasso regression, modified Hans and PC samplers. The modified Hans sampler is based on a newly defined Lasso distribution.

This vignette introduces the main functionality of the package and demonstrates how to apply the samplers to real and simulated data.

```{r}
effective_sample_size <- function(samples) {
  if (!requireNamespace("posterior", quietly = TRUE)) {
    message(
      "Package 'posterior' is required to compute effective sample size (ESS).\n",
      "Install it with install.packages('posterior') to enable ESS diagnostics."
    )
    return(NA_real_)
  }
  
  as.numeric(posterior::ess_bulk(as.numeric(samples)))
}

```

## Simulated Example
```{r}
if(dataset_name == "simulated"){

  # Simulate data
  set.seed(123)
  Ns <- 2000
  ns <- 100
  ps <- 10
  X <- matrix(rnorm(ns * ps), nrow = ns)
  beta <- c(rep(2, 3), rep(0, ps - 3))
  y <- X %*% beta + rnorm(ns)
  
  vtime_val_Hans = c()
  results_Hans <- NULL 

  # Run the modified Hans sampler
  for(i in 1:5){
    time_val <- system.time({
      res_Hans <- Modified_Hans_Gibbs(
      y = y, X = X, beta_init= rep(1,10), a1=2, b1=1, u1=2, v1=1,
      nsamples=Ns, lambda_init=1, sigma2_init=1, verbose=100, tune_lambda2 = TRUE,
      rao_blackwellization = FALSE)
  })[3]
  
  vtime_val_Hans[i] <- time_val
  
  # Initialize accumulators after first run
  if (is.null(results_Hans)) {
    results_Hans <- list(
      mBeta = res_Hans$mBeta,
      vsigma2 = res_Hans$vsigma2,
      vlambda2 = res_Hans$vlambda2
    )
  } else {
    results_Hans$mBeta    <- results_Hans$mBeta + res_Hans$mBeta
    results_Hans$vsigma2  <- results_Hans$vsigma2 + res_Hans$vsigma2
    results_Hans$vlambda2 <- results_Hans$vlambda2 + res_Hans$vlambda2
  }
  }
  
  # Take averages
  mBeta = (results_Hans$mBeta)/5
  vsigma2 = (results_Hans$vsigma2)/5
  vlambda2 = (results_Hans$vlambda2)/5
  time_val_Hans = mean(vtime_val_Hans)
  

  vESS <- c()
  for(j in 1:ps) {
    vESS[j] <- effective_sample_size(results_Hans$mBeta[200:Ns,j])
  }
  Ef_Hans = median(vESS)/time_val_Hans
  
  ESS_sigma2_Hans  = effective_sample_size(as.vector(results_Hans$vsigma2[200:Ns]))
  Ef_sigma2_Hans = ESS_sigma2_Hans/time_val_Hans
  
  ESS_lambda2_Hans  = effective_sample_size(as.vector(results_Hans$vlambda2[200:Ns]))
  Ef_lambda2_Hans = ESS_lambda2_Hans/time_val_Hans
  
  stat_vec_Hans = c(
    100*median(vESS)/Ns,
    Ef_Hans,
    100*ESS_sigma2_Hans/Ns,
    Ef_sigma2_Hans,
    100*ESS_lambda2_Hans/Ns,
    Ef_lambda2_Hans,
    time_val_Hans)
  
  name_vals = c("mix_beta", "eff_beta", "mix_sigma2", "eff_sigma2", "mix_lambda2", "eff_lambda2", "time")
  names(stat_vec_Hans) = name_vals 

  
  # ======================== PC sampler =======================================
  
  # Run the modified PC sampler
  
  vtime_val_PC = c()
  results_PC <- NULL 
  
 for(i in 1:5){
   time_val <- system.time({
     res_PC <- Modified_PC_Gibbs(
      X = X, y = y, a1=2, b1=1, u1=2, v1=1,
      nsamples=Ns, lambda_init=1, sigma2_init=1, verbose=100)
  })[3]
  
  vtime_val_PC[i] <- time_val
  
  # Initialize accumulators after first run
  if (is.null(results_PC)) {
    results_PC <- list(
      mBeta = res_PC$mBeta,
      vsigma2 = res_PC$vsigma2,
      vlambda2 = res_PC$vlambda2
    )
  } else {
    results_PC$mBeta    <- results_PC$mBeta + res_PC$mBeta
    results_PC$vsigma2  <- results_PC$vsigma2 + res_PC$vsigma2
    results_PC$vlambda2 <- results_PC$vlambda2 + res_PC$vlambda2
  }
 }
  
  # Take averages
  mBeta = (results_PC$mBeta)/5
  vsigma2 = (results_PC$vsigma2)/5
  vlambda2 = (results_PC$vlambda2)/5
  time_val_PC = mean(vtime_val_PC)
  
  # colMeans(results_PC$mBeta)
  
  vESS <- c()
  for(j in 1:ps) {
    vESS[j] <- effective_sample_size(results_PC$mBeta[200:Ns,j])
  }
  Ef_PC = median(vESS)/time_val_PC
  
  ESS_sigma2_PC  = effective_sample_size(as.vector(results_PC$vsigma2[200:Ns]))
  Ef_sigma2_PC = ESS_sigma2_PC/time_val_PC
  
  ESS_lambda2_PC  = effective_sample_size(as.vector(results_PC$vlambda2[200:Ns]))
  Ef_lambda2_PC = ESS_lambda2_PC/time_val_PC
  
  stat_vec_PC = c(
    100*median(vESS)/Ns,
    Ef_PC,
    100*ESS_sigma2_PC/Ns,
    Ef_sigma2_PC,
    100*ESS_lambda2_PC/Ns,
    Ef_lambda2_PC,
    time_val_PC)
  
  name_vals = c("mix_beta", "eff_beta", "mix_sigma2", "eff_sigma2", "mix_lambda2", "eff_lambda2", "time")
  names(stat_vec_PC) = name_vals 
}

```


## Data Sources

This vignette demonstrates the usage of the `BayesianLasso` package on several well-known datasets. To avoid redistribution issues and keep the package lightweight, none of these datasets are included directly in the package. Instead, we provide guidance on how to access them.

- **Diabetes**: Available in the \CRANpkg{lars} package and loaded using `data(diabetes, package = "lars")`.

- **Kakadu**: Available in the \CRANpkg{Ecdat} package and loaded using `data(Kakadu, package = "Ecdat")`.

- **Communities and Crime**: This dataset is available from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/183/communities+and+crime).  
  Due to licensing and redistribution considerations, the *Communities and Crime* dataset is not included in this package. Users can download the dataset manually by visiting the repository and following the preprocessing steps outlined in this vignette.


## Reproducing Table 1: Performance Comparison

This section demonstrates how to reproduce the performance comparison table from the manuscript, comparing the mixing percentages, sampling efficiencies, and runtimes for the Diabetes, Kakadu, and Communities and Crime datasets.



```{r generate-table-1}
# Load libraries
library(BayesianLasso)
# This code requires bayesreg

if (requireNamespace("monomvn", quietly = TRUE)) {
  monomvn::monomvn(...)  # use the function safely
} else {
  message("monomvn package not installed; skipping example.")
}
if (requireNamespace("bayeslm", quietly = TRUE)) {
  bayeslm::bayeslm(...)  # use the function safely
} else {
  message("bayeslm package not installed; skipping example.")
}
if (requireNamespace("rstan", quietly = TRUE)) {
  rstan::rstan(...)  # use the function safely
} else {
  message("rstan package not installed; skipping example.")
}
if (requireNamespace("bayesreg", quietly = TRUE)) {
  bayesreg::bayesreg(...)  # use the function safely
} else {
  message("bayesreg package not installed; skipping example.")
}


# Example: dataset_name <- "diabetes2"

if (dataset_name=="diabetes2") 
{
  if (!requireNamespace("lars", quietly = TRUE)) {
    install.packages("lars")
  }
  data(diabetes, package = "lars")
  
  y = diabetes$y
  x = diabetes$x
  inds = 1:ncol(x)
  
  # Normalizing and scaling the dataset by function normalize()
  norm = normalize(y,x, scale = TRUE)
  x = norm$mX
  x <- model.matrix(~.^2, data=data.frame(x=x))[,-1]
  y <- norm$vy
  
}

if (dataset_name=="Kakadu2") 
{
  if (!requireNamespace("Ecdat", quietly = TRUE)) {
    install.packages("Ecdat")
  }
  dat <- data("Kakadu", package = "Ecdat")
  
  
  # Get y vector and X matrix
  y <- as.vector(dat$income)
  x <- dat[,c(2:21,23)]  
  
  x <- model.matrix(~.^2,data=x)[,-1]
}

# Make sure Crime.csv (or comData.Rdata) is downloaded manually before running this
if (dataset_name=="Crime") 
{
  rdata_path <- system.file("extdata", "comData.Rdata", package = "BayesianLasso")
  load(rdata_path)
  # Show only if file exists
  crime_path <- "Crime.csv"
  if (file.exists(crime_path)) {
    crime_data <- read.csv(crime_path)
  } else {
    message("Please download 'communities.data' from the UCI Machine Learning Repository and convert to Crime.csv.")
  }
  
  
  # ---- Remove rows with NA while keeping X and Y aligned ----
  datXY <- na.omit(cbind(as.data.frame(X), as.data.frame(Y)))
  X2 <- as.matrix(datXY[, colnames(X), drop = FALSE])
  Y2 <- as.matrix(datXY[, colnames(Y), drop = FALSE])
  
  # ---- Drop unwanted columns safely ----
  drop_cols <- c("ownHousQrange", "rentUpperQ")
  X2 <- X2[, !colnames(X2) %in% drop_cols, drop = FALSE]
  
  # ---- Define regression inputs ----
  x <- X2
  y <- as.vector(Y2[, "murders"])
  varnames <- colnames(x)
  inds <- seq_len(ncol(x))
}


# Set prior hyperparameter constants
a1 = 1.0E-2  # Prior shape for sigma2
b1 = 1.0E-2  # Prior scale for sigma2
u1 = 1.0E-2  # Prior shape for lambda2
v1 = 1.0E-2  # Prior scale for lambda2

# Initial values for lambda2 and sigma2
lambda2_init  = 10
lambda_init = sqrt(lambda2_init)
sigma2_init = 1

# Number of samples to run the MCMC
nburn = 1000
nsamples = 5000
inds_use = (nburn + 1):nsamples
N = length(inds_use)



if(dataset_name!="simulated"){
  
  # To store elapsed time and results of the PC sampler
  vtime_val_PC = c()
  results_PC <- NULL  # to be initialized after the first run
  
  # Running the modified PC sampler 5 times and taking the average of the results across runs.
  for (i in 1:5) {
    time_val <- system.time({
      res_PC <- Modified_PC_Gibbs(
        X = x, y = y, a1, b1, u1, v1,
        nsamples,
        lambda_init = lambda_init,
        sigma2_init = sigma2_init,
        verbose = 1000
      )
    })[3]
    
    vtime_val_PC[i] <- time_val
    
    # Initialize accumulators after first run
    if (is.null(results_PC)) {
      results_PC <- list(
        mBeta = res_PC$mBeta,
        vsigma2 = res_PC$vsigma2,
        vlambda2 = res_PC$vlambda2
      )
    } else {
      results_PC$mBeta    <- results_PC$mBeta + res_PC$mBeta
      results_PC$vsigma2  <- results_PC$vsigma2 + res_PC$vsigma2
      results_PC$vlambda2 <- results_PC$vlambda2 + res_PC$vlambda2
    }
  }
  
  # Take averages
  mBeta = (res_PC$mBeta)/5
  vsigma2 = (res_PC$vsigma2)/5
  vlambda2 = (res_PC$vlambda2)/5
  time_val_PC = mean(vtime_val_PC)
  
  # Compute effective sample sizes  
  vESS <- c()
  for(j in 1:p) {
    vESS[j] <- effective_sample_size(mBeta[inds_use,j])
  }
  Ef_PC = median(vESS)/time_val_PC
  
  ESS_sigma2_PC  = effective_sample_size(as.vector(vsigma2[inds_use]))
  Ef_sigma2_PC = ESS_sigma2_PC/time_val_PC
  
  ESS_lambda2_PC  = effective_sample_size(as.vector(vlambda2[inds_use]))
  Ef_lambda2_PC = ESS_lambda2_PC/time_val_PC
  
  stat_vec_PC = c(
    100*median(vESS)/N,
    Ef_PC,
    100*ESS_sigma2_PC/N,
    Ef_sigma2_PC,
    100*ESS_lambda2_PC/N,
    Ef_lambda2_PC,
    time_val_PC)
  
  name_vals = c("mix_beta", "eff_beta", "mix_sigma2", "eff_sigma2", "mix_lambda2", "eff_lambda2", "time")
  names(stat_vec_PC) = name_vals 
  
  
  # To store elapsed time and results of the Hans sampler
  vtime_val_Hans = c()
  results_Hans <- NULL  # to be initialized after the first run
  
  # Running the modified Hans sampler 5 times and taking the average of the results across runs.
  for (i in 1:5) {
    time_val <- system.time({
      res_Hans <- Modified_Hans_Gibbs(
        y = y, X = x, beta_init = as.vector(colMeans(mBeta)),
        a1, b1, u1, v1,
        nsamples,
        lambda_init = lambda_init,
        sigma2_init = sigma2_init,
        verbose = 1000, tune_lambda2 = TRUE, rao_blackwellization = FALSE
      )
    })[3]
    
    vtime_val_Hans[i] <- time_val
    
    # Initialize accumulators after first run
    if (is.null(results_Hans)) {
      results_Hans <- list(
        mBeta = res_Hans$mBeta,
        vsigma2 = res_Hans$vsigma2,
        vlambda2 = res_Hans$vlambda2
      )
    } else {
      results_Hans$mBeta    <- results_Hans$mBeta + res_Hans$mBeta
      results_Hans$vsigma2  <- results_Hans$vsigma2 + res_Hans$vsigma2
      results_Hans$vlambda2 <- results_Hans$vlambda2 + res_Hans$vlambda2
    }
  }
  
  # Take averages
  mBeta = (results_Hans$mBeta)/5
  vsigma2 = (results_Hans$vsigma2)/5
  vlambda2 = (results_Hans$vlambda2)/5
  time_val_Hans = mean(vtime_val_Hans)
  
  # Compute effective sample sizes
  vESS <- c()
  for(j in 1:p) {
    vESS[j] <- effective_sample_size(mBeta[inds_use,j])
  }
  Ef_Hans = median(vESS)/time_val_Hans
  
  ESS_sigma2_Hans  = effective_sample_size(as.vector(vsigma2[inds_use]))
  Ef_sigma2_Hans = ESS_sigma2_Hans/time_val_Hans
  
  ESS_lambda2_Hans  = effective_sample_size(as.vector(vlambda2[inds_use]))
  Ef_lambda2_Hans = ESS_lambda2_Hans/time_val_Hans
  
  stat_vec_Hans = c(
    100*median(vESS)/N,
    Ef_Hans,
    100*ESS_sigma2_Hans/N,
    Ef_sigma2_Hans,
    100*ESS_lambda2_Hans/N,
    Ef_lambda2_Hans,
    time_val_Hans)
  
  name_vals = c("mix_beta", "eff_beta", "mix_sigma2", "eff_sigma2", "mix_lambda2", "eff_lambda2", "time")
  names(stat_vec_Hans) = name_vals 
  
  
  if(dataset_name !="Kakadu2"){
    res_monomvn <- NULL
    
    if (requireNamespace("monomvn", quietly = TRUE)) {
      
      
      res_monomvn = benchmark_blasso_monomvn(y, x,
                                             lambda_init, sigma2_init, a, b, u, v, 
                                             nburn, nsamples, trials=5, 
                                             beta_inds=NA) 
      
      stat_vec_monomvn <- colMeans(res_monomvn$mStat)
      
      
    }else {
      stop("Skipping monomvn benchmark: package 'monomvn' is not installed.")
    }
  }
  
  
  # ======================= Run bayeslm ============================
  
  res_bayeslm <- NULL
  
  if (requireNamespace("bayeslm", quietly = TRUE)) {
    
    res_bayeslm = benchmark_blasso_bayeslm(y, x,
                                           lambda_init, sigma2_init, a, b, u, v, 
                                           nburn, nsamples, trials=5, 
                                           beta_inds=NA) 
    
    stat_vec_bayeslm <- colMeans(res_bayeslm$mStat)
    
    
  }else {
    stop("Skipping bayeslm benchmark: package 'bayeslm' is not installed.")
  }
  
  
  # ======================= Run rstan ============================
  
  res_rstan <- NULL
  
  if (requireNamespace("rstan", quietly = TRUE)) {
    
    res_rstan = benchmark_blasso_rstan(y, x,
                                       lambda_init, sigma2_init, a, b, u, v, 
                                       nburn, nsamples, trials=5, 
                                       beta_inds=NA)
    
    stat_vec_rstan <- colMeans(res_rstan$mStat)
    
    
  }else {
    stop("Skipping rstan benchmark: package 'rstan' is not installed.")
  }
  
  
  # ======================= Run bayesreg ============================
  
  
  res_bayesreg <- NULL
  
  if (requireNamespace("bayesreg", quietly = TRUE)) {
    
    res_bayesreg = benchmark_blasso_bayesreg(y, x,
                                             lambda_init, sigma2_init, a, b, u, v, 
                                             nburn, nsamples, trials=5, 
                                             beta_inds=NA) 
    
    stat_vec_bayesreg <- colMeans(res_bayesreg$mStat)
    
    
  }else {
    stop("Skipping bayesreg benchmark: package 'bayesreg' is not installed.")
  }
  
}

```

## Convergence diagnostics

Assessing convergence of the MCMC chains is an important step in Bayesian analysis. 
The package provides helper functions such as `mcmc_stats()` and `mcmc_diagnostics()` 
to summarize mixing behavior and visualize the behavior of the chains. The function 
`mcmc_stats()` computes numerical summaries of the MCMC output, while 
`mcmc_diagnostics()` produces basic diagnostic plots such as trace and autocorrelation 
plots for selected parameters. These tools allow users to assess mixing and stability 
of the chains in practice. Posterior draws can also be passed to packages such as 
`coda`, `posterior`, or `bayesplot` for additional diagnostics.





